pdp: An R Package for Constructing Partial Dependence Plots

### 2021-08-13 rjournal/Untitled/greenwell.Rmd

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Front matter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{pdp: An R Package for Constructing Partial Dependence Plots}
\author{by Brandon M. Greenwell}

\maketitle

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Constructing PDPs in R
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructing PDPs in R}
\label{sec:boston}

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph; hence, the same is true when using \code{partial} with \code{plot = TRUE}.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}). This is demonstrated in multiple examples throughout this paper.

@@

For illustration, we will use a corrected version of the Boston housing data analyzed in \citet{harrison-1978-hedonic}; the data are available in the \pkg{pdp} package (see \code{?pdp::boston} for details). We begin by loading the data and fitting a random forest with default tuning parameters and 500 trees:
\begin{example}
data(boston, package = "pdp")  # load the (corrected) Boston housing data
library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # Figure 1
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
  \label{fig:plotmo_vs_partial}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Single predictor PDPs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Single predictor PDPs}

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor". For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
\begin{example}
library(pdp)  # for partial, plotPartial, and grid.arrange functions
partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
\end{example}
The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.

For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame. This is illustrated in the example below which increases the line width, adds a LOESS smooth, and customizes the $y$-axis label. The result is displayed in the right side of Figure~\ref{fig:pd_lstat}. \strong{Note:} to encourage writing more readable code, the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded.
\begin{example}
# Figure 2 (right)
boston.rf %>%  # the %>% operator is read as "and then"
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multi-predictor PDPs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multi-predictor PDPs}
\label{sec:multi}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Avoiding extrapolation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Avoiding extrapolation}
\label{sec:extrapolation}

It is not wise to draw conclusions from PDPs in regions outside the area of the training data. Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls. Rug displays are one-dimensional plots added to the axes. Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes. The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
\begin{example}
# Figure 4 (left)
partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
\end{example}

In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted. Over interpreting the PDP outside of this region is considered extrapolation and is ill-advised. The right display in Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
# Figure 4 (right)
partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_extrap}
  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
  \label{fig:partial_extrap}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Addressing computational concerns
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Addressing computational concerns}
\label{sec:computational}

Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; \code{grid.arrange} is used to display all three PDPs on the same graph, side by side. The results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
# Figure 5
grid.arrange(
  partial(boston.rf, "rm", plot = TRUE),
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_manual}
  \caption{Partial dependence of \code{cmedv} on \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a reduced grid size. \textit{Right}: Using a user-specified grid.}
  \label{fig:partial_manual}
\end{widefigure}

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops. This makes it easy to request progress bars (e.g., \code{progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].

To illustrate, we will use the Los Angeles ozone pollution data described in \citet{breiman-1985-estimating}. The data contain daily measurements of ozone concentration (\code{ozone}) along with eight meteorological quantities for 330 days in the Los Angeles basin in 1976.\footnote{The data are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data}. Details, including variable information, are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.info}.} The following code chunk loads the data into R:
\begin{example}
ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
                         "datasets/LAozone.data"), header = TRUE)
\end{example}

Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
\begin{example}
library(earth)  # for earth function (i.e., MARS algorithm)
ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
summary(ozone.mars)
\end{example}
The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
\begin{itemize}
  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
\end{itemize}
To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load the parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)  # register the parallel backend
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The last three lines of code add a label to the colorkey. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice

# Add a label to the colorkey
lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
grid::grid.text("ozone", x = 0.2, y = 1.05, hjust = 0.5, vjust = 1)
lattice::trellis.unfocus()
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}. Since \code{dpg} is continuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}). Hence, it will be more effective to use categorical variables to define the panels in higher dimensional displays when possible.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Classification problems
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Classification problems}
\label{sec:classification}

Traditionally, for classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp. 369---370]{hastie-elements-2009}. Suppose the response is categorical with $K$ levels, then for each class we compute
\begin{equation}
\label{eqn:avg-logit}
f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{k = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
\end{equation}
where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.

To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package. The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris: setosa, versicolor, and virginica. We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)  # for svm function
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                cost = 0.25, probability = TRUE)
\end{example}
\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

# Figure 7
library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
  geom_tile() +
  geom_contour(color = "white", alpha = 0.5) +
  scale_fill_distiller(name = "Centered\nlogit", palette = "Spectral") +
  theme_bw() +
  facet_grid(~ Species)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} for the iris data.}
  \label{fig:partial_iris}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% User-defined prediction functions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{User-defined prediction functions}
\label{sec:prediction}

PDPs are essentially just averaged predictions; this is apparent from step 1. (c) in Algorithm~\ref{alg:pdp}. Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong heterogeneity can conceal the complexity of the modeled relationship between the response and predictors of interest. This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.

With \code{partial} it is possible to replace the mean in step 1. (c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either a single prediction or a vector of predictions. Returning the mean of all the predictions will result in the traditional PDP. Returning a vector of predictions (i.e., one for each observation) will result in a set of ICE curves. The examples below illustrate.

Using the \code{pred.fun} argument, it is possible to obtain PDPs for classification problems on the probability scale. We just need to write a function that computes the predicted class probability of interest averaged across all observations. The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:classification} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The following chunk of code uses \code{pred.prob} to obtain PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale. The results are displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
# PDPs for Petal.Width and Petal.Length on the probability scale
pdp.pw <- partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob,
                  plot = TRUE)
pdp.pl <- partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob,
                  plot = TRUE)
pdp.pw.pl <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                     pred.fun = pred.prob, plot = TRUE)

# Figure 8
grid.arrange(pdp.pw, pdp.pl, pdp.pw.pl, ncol = 3)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probability of belonging to the setosa species.}
  \label{fig:partial_iris_prob}
\end{figure}

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata} (i.e., just remove the call to \code{mean} in \code{pred.fun}). The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}. The result is displayed in Figure~\ref{fig:ice_boston}. \strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in the following code chunk, there will be one curve for each observation in \code{boston}.
\begin{example}
# Use partial to obtain ICE curves
pred.ice <- function(object, newdata) predict(object, newdata)
rm.ice <- partial(boston.rf, pred.var = "rm", pred.fun = pred.ice)

# Figure 9
plotPartial(rm.ice, rug = TRUE, train = boston, alpha = 0.3)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{ice_boston}
  \caption{ICE curves depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. Each curve corresponds to a different observation.}
  \label{fig:ice_boston}
\end{figure}

The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship). Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}. Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg. 130]{fitzmaurice-2011-applied} for longitudinal data). This is shown below.
\begin{example}
# Post-process rm.ice to obtain c-ICE curves
library(dplyr)  # for group_by and mutate functions
rm.ice <- rm.ice %>%
  group_by(yhat.id) %>%  # perform next operation within each yhat.id
  mutate(yhat.centered = yhat - first(yhat))  # so each curve starts at yhat = 0
\end{example}

Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot. This is easily accomplished using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together. The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example. The results are displayed in Figure~\ref{fig:ice_cice_boston}.
\begin{example}
# ICE curves with their average
p1 <- ggplot(rm.ice, aes(rm, yhat)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

# c-ICE curves with their average
p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

# Figure 10
grid.arrange(p1, p2, ncol = 2)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. \textit{Left}: Uncentered (here the red curve is just the traditional PDP). \textit{Right}: Centered.}
  \label{fig:ice_cice_boston}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Using partial with xgboost
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Using partial with the XGBoost library}
\label{sec:xgboost}

>>>>

---
title: "pdp: An R Package for Constructing Partial Dependence Plots"
author:
  # see ?rjournal_article for more information
  - name: Author One
    affiliation: Affiliation
    address:
    - line 1
    - line 2
    url: https://journal.r-project.org
    orcid: 0000-0002-9079-593X
    email:  author1@work
abstract: >
  Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
preamble: |
  \usepackage[utf8]{inputenc}
  \usepackage[T1]{fontenc}
  \usepackage{RJournal}
  \usepackage{amsmath,amssymb,array}
  \usepackage{booktabs}
  \usepackage{algorithm}

# per R journal requirement, the bib filename should be the same as the output 
# tex file. Don't forget to rename the bib file and change this example value.
bibliography: greenwell.bib

output: rticles::rjournal_article
---

```{r setup, include=FALSE}
library(magrittr)
library(pdp)

# Set global knitr chunk options
knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "100%",
  fig.align = "center",
  fig.pos = "!htb",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

**Note:** This is an updated version of the original R Journal article.

@@

Several additional packages have been developed in recent years that provide similar functionality to \pkg{pdp}, such as the \CRANpkg{iml} package \citep{iml-pkg}.

@@

<!----------------------------------------------------------------------------->
\section{Constructing PDPs in R}
<!----------------------------------------------------------------------------->

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph; hence, the same is true when using \code{partial} with \code{plot = TRUE}.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}. For example, the \code{grid.arrange} function from \pkg{gridExtra} makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}). This is demonstrated in multiple examples throughout this paper.

@@

For illustration, we will use a corrected version of the Boston housing data analyzed in \citet{harrison-1978-hedonic}; the data are available in the \pkg{pdp} package (see \code{?pdp::boston} for details). We begin by loading the data and fitting a random forest with default tuning parameters and 500 trees:

```{r boston-vip, fig.cap="Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees."}
library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions

data(boston, package = "pdp")  # load the (corrected) Boston housing data

# Fit a default random forest to the Boston housing data
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # Figure 1
```

The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:boston-vip}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

<!----------------------------------------------------------------------------->
\subsection{Single predictor PDPs}
<!----------------------------------------------------------------------------->

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor". For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:

```{r boston-lstat-pd, eval=FALSE}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
```

The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):

```{r boston-lstat-pd-plot, eval=FALSE}
library(pdp)  # for partial, plotPartial, and grid.arrange functions
partial(boston.rf, pred.var = "lstat", plot = TRUE)
```

The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package by default to produce all of its displays.

```{r boston-lstat-pd-plot-ggplot2, eval=FALSE}
partial(boston.rf, pred.var = "lstat", plot = TRUE, plot.engine = "ggplot2")
```

For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial()} (for \pkg{lattice} graphics) or \code{autoplot()} (for \pkg{ggplot2} graphics) functions on the resulting data frame. This is illustrated in the example below which increases the line width, adds a LOESS smooth, and customizes the $y$-axis label. The result is displayed in the right side of Figure~\ref{fig:pd_lstat}. 
```{r boston-lstat-pd-plotPartial, eval=FALSE}
library(magrittr)

boston.rf %>%  # the %>% operator is read as "and then"
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
```

```{r pd-lstat, fig.cap="Partial dependence of \\code{cmedv} on \code{lstat} based on a random forest. \\textit{Left}: Default plot. \\textit{Right}: Customized plot obtained using the \\code{plotPartial()} function."}
p1 <- partial(boston.rf, pred.var = "lstat", plot = TRUE, plot.engine = "ggplot2")
p2 <- boston.rf %>%  # the %>% operator is read as "and then"
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
gridExtra::grid.arrange(p1, p2, nrow = 1)
```

```{r boston-lstat-pdps, fig.cap="Partial dependence of \code{cmedv} on \code{lstat} based on a random forest using both \\pkg{lattice}- and \pkg{ggplot2}-based graphics with various options."}
library(ggplot2)
library(gridExtra)

p1 <- partial(boston.rf, pred.var = "lstat", plot = TRUE, plot.engine = "ggplot2")
p2 <- boston.rf %>%  # the %>% operator is read as "and then"
  partial(pred.var = "lstat") %>%
  autoplot(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))

# Display plots in a grid
grid.arrange(p1, p2, nrow = 1)
```

<!----------------------------------------------------------------------------->
\subsection{Multi-predictor PDPs}
<!----------------------------------------------------------------------------->

@@
